On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-03 manuscripts/PNAS/review/point_by_point_response.txt

-- I think the biggest potential issue for publication in PNAS is that the significance of the result needs more discussion. Specifically, the authors could address why going from the noise-free case to the noisy case is not incremental. PNAS specifically targets submissions that go beyond what could be published in a more specialized journal. Could the authors, for example, contrast the proof technique between the noisy and noise-free cases? Or, perhaps, the practical implications? 
-----------------

RESPONSE:

With all due respect to the reviewer, we believe the most compelling practical implication of a noisy result over a noiseless result is plainly evident: never has there been an experiment without noise and an unstable model is therefore useless in practice. We assume the reviewer must instead be concerned with what additional practical “take-away” message our manuscript contains. In brief, it is this: there exists a procedure by which one can determine if one’s proposed solution to Prob. 1 or Prob. 2 is indeed unique. We have updated the manuscript with an explicit statement of this fact.

The reviewer also raises the concern that, regardless of practical implications, our results may amount to an incremental advance over those of [Hillar & Sommer (2015)] requiring only a straight-forward extension of their approach. We disagree with this assessment on both counts: our main result (Thm. 1) goes far beyond a simple extension of that in [Hillar & Sommer (2015)] to the noisy case, and our method of proof includes a significant deviation from their approach.

@@

3) No spark condition requirement for the generating matrix A: One of the mathematically significant contributions made by [Hillar & Sommer (2015)] was to forgo the constraint that the recovery matrix B also satisfy the spark condition, in contrast to all previously known uniqueness results which (either explicitly or implicitly) made this assumption. Our proof is powerful enough to show that, in fact, even the matrix A need not satisfy the spark condition to be identifiable! Rather, it need only be injective on the union of subspaces with supports that form sets in a hypergraph satisfying the SIP. We had omitted this point in our original submission of the manuscript to keep things simple, but have decided to include this interesting fact in our resubmission given that a recently approved NSF grant _____ makes use of this fact in its description of methods. This also required us to redefine the restricted matrix lower bound L to be in terms of a hypergraph, which is an interesting object for further study in its own right.

[4) Implications for the computational complexity of the dictionary learning problem: If the size of the support set of reconstructing codes is polynomial in m,k then the pigeonholing argument in the proof of Thm. 1 requires only a polynomial number of samples distributed over a polynomial number of supports; thus N is polynomial in m, k. [Chris? How does this have implications..?] This point was only hinted at in the Discussion section of our original submission, but we have included it in the updated manuscript to make clear the power of our approach over that taken in [Hillar & Sommer (2015)].

We also wish to reiterate that our solution to Prob. 2, that of most interest to practitioners of dictionary learning methods, is the first of its kind in both the noise-free and noisy domains. We recognize that this was not clearly communicated in the Discussion section of our submitted manuscript, which begins: “In this note, we generalized the approach of (18) to prove the stability of unique solutions to Probs. 1 and 2 while significantly reducing the known sample complexity.” We have revised our re-submitted manuscript to make it clear that a solution to Problem 2 has never been supplied before in the literature until now [right?].

@@

The paper's writing and composition is clear (with an exception detailed below), concise, and relevant to the topic. The analytical work appears to be correct to the best of my understanding, and the paper's combined use of several results in the literature to extend an existing dictionary learning uniqueness condition to a new dictionary learning stability condition appears to be original.

Some questions that remain and may be worth discussing deal with the practical impact of these results.

When is it possible to determine (numerically) if enough data has been gathered to obtain an accurate estimate of the underlying sparsity dictionary? 
-----------------

RESPONSE: A statement of this very general nature requires that we calculate the probability that all of the conditions of Thm. 1 are satisfied by random data. We have provided an ‘almost certain’ guarantee of this flavor in Cor 3, wherein the sparse vector supports are constrained to form hypergraph satisfying the SIP. It would be interesting to calculate furthermore the probability that random supports form a hypergraph satisfying the SIP. We think we have cracked open a nice set of interesting combinatorial problems (with practical implications!) to be solved by the community.

@@

RESPONSE: Yes, there is indeed. While all existing uniqueness results require a sufficient number of samples be drawn from *each* of {m \choose k} possible supports, we have shown that in fact it is enough that a sufficient number of samples be drawn from every support in a subset of drastically smaller size, provided this subset forms a regular hypergraph satisfying the singleton intersection property. As was pointed out in our manuscript, a support set of size m satisfying these criteria can always be constructed for any k < m (e.g. take the consecutive intervals of length k in some cyclic order on [m]), and in certain cases such a set can be constructed from as few as 2\sqrt(m) supports (e.g. when m = k^2, take the supports to be the rows and columns of [m] arranged in a square grid).

-----------------
In these terms, it is also unclear to me if a theoretical extension of the uniqueness results to the stability results merit publication in PNAS given the "top 10%" and "sufficient general interest" requirements for publication. It definitely is deserving of publication in a specialty journal.

The work presented is analytical and does not rely on data validation. 
-----------------

RESPONSE: Both reviewers have raised this concern and we again thank the editor for the opportunity to address it. It seems both reviewers haven’t noticed the several ways in which our results generalize the modern results of [Hillar & Sommer (2015)] beyond a straight-forward extension to the noisy case, not to mention our solution to the optimization Prob. 2, which is the first of its kind and of most interest to practitioners. We have already discussed these points in detail in our response to the first reviewer’s identical comment.

@@

-----------------
In the paragraph "Since f is an analytic function..." (page 3) it is argued that a specific construction of a suitable matrix implies that "almost every real nxm matrix... satisfies Eq. (2)". This appears to overstate the result, in my opinion - can the authors elaborate on why "almost every real matrix" can be obtained from this construction based on a Vandermonde matrix with additional zero rows? 
-----------------

RESPONSE: The point we make here is not that almost every real matrix satisfying Eq. (2) can be obtained from the construction based on a Vandermonde matrix which we provide; rather, the point is that almost every matrix satisfies Eq. (2) because at least one matrix does (e.g. our construction) and therefore, since f is analytic, the set of all those that don’t satisfy the condition has measure zero. We have updated the manuscript to clarify this argument for our readers.

### 2017-08-05 manuscripts/PNAS/chazthm_pnas.tex
L1-norm extension, references omitted

\acknow{We thank Friedrich Sommer and Darren Rhea for early thoughts, and Ian Morris for posting \eqref{eqdim} online.}
\showacknow % Display the acknowledgments section

\bibliography{chazthm_pnas}

@@

%\pnasbreak

>>>>

%\acknow{We thank Friedrich Sommer and Darren Rhea for early thoughts, and Ian Morris for posting \eqref{eqdim} online.}
%\showacknow % Display the acknowledgments section

%\bibliography{chazthm_pnas}

@@

\clearpage

\section{$\ell_1$-norm Extension}

\begin{problem}\label{OptimizationProblemL1}
Find a matrix $\mathbf{B} \in \mathbb{R}^m$ and vectors \mbox{$\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$} that solve:
\begin{align}\label{minsum}
\min \sum_{i = 1}^N \|\mathbf{\bar x}_i\|_1 \ \ 
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \eta, \ \text{for all $i$}.
\end{align}
\end{problem}

\begin{corollary}\label{SLCopt}
Fix $\alpha > 0$ and suppose all of the assumptions of Thm.\ref{DeterministicUniquenessTheorem} hold with the following modifications:
\begin{enumerate}
\item The nonzero elements of all $k$-sparse $\mathbf{x}_i$ are bounded above by $\beta$, where $\beta < \frac{\alpha(k+1)}{k}$.
\item More than $(k-1){\bar m \choose k} + q$ vectors $\mathbf{x}_i$ are supported in each $S \in \mathcal{H}$, where
\[ q = \left[\frac{\alpha}{\beta}(k+1) - k\right]^{-1}k|\mathcal{H}|(k-1){\bar m \choose k-1}. \]
\end{enumerate}
Then all solutions to Prob.~\ref{OptimizationProblemL1} for which the nonzero elements of all $\mathbf{\bar x}_i$ are bounded below by $\alpha$ necessarily satisfy the implications \eqref{Cstable} and \eqref{b-PDa} of Thm.~\ref{DeterministicUniquenessTheorem}.
\end{corollary}

Next, observe that the assumptions of the lemma imply:
\begin{align*}
\alpha \sum_i \|\mathbf{\bar x}_i\|_0 \leq \sum_i \|\mathbf{\bar x}_i\|_1 \leq \sum_i \|\mathbf{x}_i\|_1 \leq \beta \sum_i \|\mathbf{\bar x}_i\|_0.
\end{align*}

We now apply these facts to bound the number of $k$-sparse $\mathbf{\bar x}_i$. Let $n_p$ be the number of $\mathbf{\bar x}_i$ with $\|\mathbf{\bar x}_i\|_0 = p$. Since the $\mathbf{x}_i$ are all $k$-sparse, by \eqref{minsum} we have:
\begin{align*}
k \sum_{p = 0}^{\bar m} n_p = k N \geq \sum_{i=0}^N \|\mathbf{x}_i\|_0 \geq \frac{\alpha}{\beta} \sum_{i=0}^N \|\mathbf{\bar x}_i\|_0 = \frac{\alpha}{\beta} \sum_{p=0}^{\bar m} p n_p.
\end{align*}
Hence, 
\begin{align*}
\left[ \frac{\alpha(k+1)}{\beta} - k\right] \sum_{p = k+1}^{\bar m} n_p 
&\leq \sum_{p = k+1}^{\bar m} (\frac{\alpha}{\beta} p-k) n_p
\leq \sum_{p = 0}^k (k - \frac{\alpha}{\beta} p)n_p \\
&\leq k \sum_{p = 0}^{k-1} n_p 
\leq k|\mathcal{H}|(k-1){ \bar m \choose k-1}
\end{align*}

Therefore, no more than $q$ vectors $\mathbf{\bar x}_i$ are \emph{not} $k$-sparse. Since for every $S \in \mathcal{H}$ there are over $(k-1){\bar m \choose k} + q$ vectors $\mathbf{x}_i$ supported there, it follows that more than $(k-1){\bar m \choose k}$ of them have corresponding $\mathbf{\bar x}_i$ that are $k$-sparse. The result now follows directly from Thm.~\ref{DeterministicUniquenessCorollary}.
\end{proof}
